Compositional Quantitative Resource Analysis of C Programs

### 2014-06-21 paper/main.tex
some updates to the new logical derivation

Since programs can be composed we must enrich the evaluation judgement
$(p, \state) \Downarrow_{n}$ to talk about the state resulting from
the program execution.  We write $(p, \state) \Downarrow_n \state'$ to
say that $\state'$ is the program state after the execution of $p$ in
starting state $\sigma$.  We now use two potential functions, one that
applies before the execution, and one that applies after.  The two
functions must respect the relation $\Phi(\state) \ge n +
\Phi'(\state')$ for all states $\state$ and $\state'$ such that $(p,
\state) \Downarrow_n \state'$.  Intuitively, $\Phi(\state)$ must
provide enough \emph{potential} for, both, paying for resource cost of
the computation and paying for the potential $\Phi'(\state')$ on the
resulting state $\state'$. That way, if $(\state, p_1) \Downarrow_n
\state'$ and $(\state', p_2) \Downarrow_m \state''$, we get
$\Phi(\state) \ge n + \Phi'(\state')$ and $\Phi'(\state') \ge m +
\Phi''(\state'')$.  This can be composed to $\Phi(\state) \ge (n + m)
+ \Phi''(\state'')$.  Note that the initial potential function $\Phi$
provides an upper bound on the resource consumption of the whole
program.  What we have observed is that, if we define $\htriple{\Phi}
{p}{\Phi'}$ to mean
$$
\forall \state, n, \state' : (\state, p) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state')
$$
then we get the following familiar looking rule.
$$
\RuleNolabel
{\htriple{\Phi}{p_1}{\Phi'} \\ \htriple{\Phi'}{p_2}{\Phi''}}
{\htriple{\Phi} {p_1; p_2}  {\Phi''}}
$$
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect systems
rules.  This rules enable to interactively reason about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

@@

\begin{figure}
\begin{lstlisting}
${\color{blue}%
  \{ \#_1(a) + 2k \}}$
while (k > 0) {
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
  while (a[x] == 1 && x < N) {
    ${\color{blue}%
      \{ (a[x] = 1) + \#_1(a) + 2k \}}$
    a[x]=0;
    ${\color{blue}%
      \{ (a[x] = 0) + \#_1(a) + 2k + 1 \}}$
    tick(1);
    ${\color{blue}%
      \{ \#_1(a) + 2k \}}$
    x++; }
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
  if (x < N) {
    ${\color{blue}%
      \{ \#_1(a) + 1 + 2(k-1) + 1 \}}$
    a[x]=1;
    ${\color{blue}%
      \{ \#_1(a) + 2(k-1) + 1 \}}$
    tick(1); }
  ${\color{blue}%
    \{ \#_1(a) + 2(k-1) \}}$
  k--;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
} ${\color{blue} \{ 0 \}}$

\end{lstlisting}
\caption{Exemple derivation using amortized reasoning, we noted
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$, we use the
  tick metric that assigns cost $n$ to \lstinline{tick(n)} statments and
  0 to other operations.
  }
\label{fig:xmplinc}
\end{figure}

@@

\and \Rule{Q:Set}
{ \forall u,~q_{yu} = q'_{xu} + q'_{yu}
\\ q'_{xy} \in \mathbb Q_0^+
}
{ \Gamma[x/y]; Q + M_s \vdash x \gets y \dashv Q'; \Gamma }

@@

\and \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |x - y| \}
\\ \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |x - y| \le u_{xy} \}
%\\ q'_0 = q_0 + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i
\\ \forall i \in \mathcal U,~ q'_i \ge q_i - p_i
\\ \forall i \in \mathcal L,~ q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U \cup \mathcal L,~ q'_i \ge q_i
}
{ Q' + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i  \succeq_\Gamma Q }

@@

The rules of~\pref{fig:proof} are designed to enforce a semantic
soundness similar to the one exposed in the overview section.  To
do this, the rules impose contraints on $Q$ and $Q'$.  To make the
presentation lighter, we use the following  set of conventions.
First, the context $Q'$ noted $Q \pm n$ is defined by $q'_0 = q_0 \pm n$
and $\forall i \neq 0,~q'_i = q_i$.  Second, for any index $i$ such that
the relation between $q_i$ and $q'_i$ is not specified as a side condition
of a rule, we assume that $q_i = q'_i$.  Third, we implicitely assume
all the side conditions enforcing that the coefficients $(q_i)_{i \in I}$ are
non-negative numbers. Finally, when we write $q_x$ for $q_{x0}$ where
0 is a program variable symbolizing the constant 0.

@@

Syntactically $\Gamma$ is a list of program conditions.  Because of the
rules {\sc Q:While} and {\sc Q:If}, we require the program conditions
to be closed by negation.  Semantically we define the validity
$H \models \Gamma$ of $\Gamma = C_1, \dots, C_n$ on a given
heap $H$ to be $\bigwedge_i \llbracket C_i \rrbracket_H = \top$.
%
The logical state
is maintained using regular Hoare style rules.
The semantic entailment on logical states is noted $\Gamma \models \Delta$,
its meaning is that $\forall H,~ H\models\Gamma \implies H\models\Delta$.

@@

\begin{lemma}[Relax]
If $Q' \succeq_\Gamma Q$, for any heap $H$ such that $H \models \Gamma$,
we have $\Phi_H(Q') \ge \Phi_H(Q)$.
\end{lemma}
\begin{proof}
We observe the following inequations.
\begin{align*}
\Phi_H(Q') &= \sum_i q'_i f_i(H) \\
&\ge q_0 + \sum_{i\in\mathcal U} u_i p_i - \sum_{i\in\mathcal L} l_i p_i
  + \sum_{i\in\mathcal U} (q_i - p_i) f_i(H) + \sum_{i\in\mathcal L} (q_i + p_i) f_i(H)
  + \sum_{i\not\in \mathcal U \cup \mathcal L} q_i \\
 &\ge \Phi_H(Q)
  + \sum_{i\in\mathcal U} (u_i - f_i(H)) p_i
  + \sum_{i\in\mathcal L} (f_i(H) - l_i) p_i \\
 &\ge \Phi_H(Q).
\end{align*}
Since $H \models \Gamma$, using the transitivity of $\models$ and
the definition of $\mathcal U$ and $\mathcal L$ we get
$\forall i\in\mathcal L,~ l_i \le f_i(H)$ and
$\forall i\in\mathcal U,~ f_i(H)\le u_i$.
These two inequalities justify the last step.
\end{proof}

@@

\item Case {\sc S:Seq}.
  By transitivity and induction hypothesis, it is
  easy to get $H'' \models \Gamma''$.  Now we prove
  the quantitative part of the theorem.  We have
  $(q, q') = M_s^1 \cdot (q_1, q'_1) \cdot M_s^2 \cdot (q_2, q'_2)$.
  We assume $\forall i \in \{1, 2\},~M_s^i \ge 0$, the other cases
  are similar.
  \begin{itemize}[topsep=0pt]
  \item
    The first case is when $q'_1 > M_s^2 + q_2$ then $q = M_s^1 + q_1$
    and since $q_1 \le \Phi_H(Q - M_s^1)$ (by induction) we have
    $q \le \Phi_H(Q)$.
  \item
    The second case is when $q'_1 \le M_s^2 + q_2$ then $q = M_s^1 + q_1 + M_s^2 + q_2 - q'_1$. \\
    Since $\Phi_H(Q - M_s^1) - \Phi_{H'}(Q') \ge q_1 - q'_1$
    and $q_2 \le \Phi_{H'}(Q' - M_s^2)$ we have:
    \begin{align*}
    q &\le M_s^1 + \Phi_H(Q - M_s^1) - \Phi_{H'}(Q') + M_s^2 + q_2 \\
      &= \Phi_H(Q) + (q_2 - \Phi_{H'}(Q' - M_s^2)) \le \Phi_H(Q).
   \end{align*}
  \end{itemize}
  We obtain $\Phi_H(Q) - \Phi_{H''}(Q'') \ge q - q'$ by simply chaining
  the two induction hypothesis and using the fact that $q - q' =
  (q_1 + M_s^1 - q'_1) + (q_2 + M_s^2 - q'_2)$.

@@

\item Case {\sc S:WhileStep}.
  Here again we assume $\forall i \in \{1, 2, 3\},~M_w^i \ge 0$.
  \begin{itemize}[topsep=0pt]
  \item
    If $q'_1 > M_w^2 + q_2$ then $q = M_w^1 + q_1$ and by induction
    we get $q_1 \le \Phi_H(Q - M_w^1)$, so $q \le \Phi_H(Q)$.
  \item
    If $q'_1 \le M_w^2 + q_2$ then $q = M_w^1 + q_1 + M_w^2 + q_2 - q'_1$.
    And we proceed like in the sequence case.
  \end{itemize}
  We obtain $\Phi_H(Q) - \Phi_{H''}(Q - M_w^3) \ge q - q'$ using the identity
  $q - q' = (M_w^1 + M_w^2 + q_1 - q'_1) + q_2 - q'_2$ and the two
  induction hypothesis
  \begin{itemize}[topsep=0pt]
  \item $\Phi_H(Q - M_w^1) - \Phi_{H'}(Q + M_w^2) \ge q_1 - q'_1$ and
  \item $\Phi_{H'}(Q) - \Phi_{H''}(Q - M_w^3) \ge q_2 - q'_2.$
  \end{itemize}

@@

\paragraph{Decision procedure.}
To decide logical entailment $\Gamma \models \Delta$ we use
Presburger's decision procedure.  But what this procedure gives us
is in fact only an answer the the question: Is there any heap $H$
such that $H \models \Gamma$.  While this sentence is existential
the question $\Gamma \models \Delta$ is universal.  To use the
decision procedure, we use the well known trick of double negation.
\begin{align*}
\Gamma \models C
&\Leftrightarrow \neg\neg (\Gamma \models C) \\
&\Leftrightarrow \neg\neg (\forall H, H\models\Gamma \implies H\models C) \\
&\Leftrightarrow \neg \exists H, H\models\Gamma \land \neg H\models C \\
&\Leftrightarrow \neg \exists H, H\models\Gamma \land H\models \neg C \\
&\Leftrightarrow \neg \exists H, H\models (\Gamma, \neg C)
\end{align*}
Thus, to know if $\Gamma \models C$ we ask the decision
procedure if $\Gamma, \neg C$ is satisfiable and negate the result.
When we want to know about $\Gamma \models \Delta$ we simply ask
one question for each of the conjuncts of $\Delta$.

>>>>

Since programs can be composed we must enrich the evaluation judgement
$(p, \state) \Downarrow_{n}$ to talk about the state resulting from
the program execution.  We write $(p, \state) \Downarrow_n \state'$ to
say that $\state'$ is the program state after the execution of $p$ in
starting state $\sigma$.  We now use two potential functions, one that
applies before the execution, and one that applies after.  The two
functions must respect the relation $\Phi(\state) \ge n +
\Phi'(\state')$ for all states $\state$ and $\state'$ such that $(p,
\state) \Downarrow_n \state'$.  Intuitively, $\Phi(\state)$ must
provide enough \emph{potential} for, both, paying for resource cost of
the computation and paying for the potential $\Phi'(\state')$ on the
resulting state $\state'$. That way, if $(\state, p_1) \Downarrow_n
\state'$ and $(\state', p_2) \Downarrow_m \state''$, we get
$\Phi(\state) \ge n + \Phi'(\state')$ and $\Phi'(\state') \ge m +
\Phi''(\state'')$.  This can be composed to $\Phi(\state) \ge (n + m)
+ \Phi''(\state'')$.  Note that the initial potential function $\Phi$
provides an upper bound on the resource consumption of the whole
program.  What we have observed is that, if we define $\htriple{\Phi}
{p}{\Phi'}$ to mean
$$
\forall \state\, n\, \state'.\, (\state, p) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state')
$$
then we get the following familiar looking rule.
$$
\RuleNolabel
{\htriple{\Phi}{p_1}{\Phi'} \\ \htriple{\Phi'}{p_2}{\Phi''}}
{\htriple{\Phi} {p_1; p_2}  {\Phi''}}
$$
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect systems
rules.  This rules enable to interactively reason about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

@@

\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}
${\color{blue}%
  \{ \#_1(a) + 2k \}}$
while (k > 0) {
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
  while (x < N && a[x] == 1) {
    ${\color{blue}%
      \{ (a[x] = 1) + \#_1(a) + 2k \}}$
    a[x]=0;
    ${\color{blue}%
      \{ \#_1(a) + 2k + 1 \}}$
    tick(1);
    ${\color{blue}%
      \{ \#_1(a) + 2k \}}$
    x++; }
  ${\color{blue}%
    \{ (x \ge N \lor a[x] = 0) + \#_1(a) + 2k \}}$
  if (x < N) {
    ${\color{blue}%
      \{ (a[x] = 0) + \#_1(a) + 1 + 2(k-1) + 1 \}}$
    a[x]=1;
    ${\color{blue}%
      \{ \#_1(a) + 2(k-1) + 1 \}}$
    tick(1); }
  ${\color{blue}%
    \{ \#_1(a) + 2(k-1) \}}$
  k--;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
} ${\color{blue} \{ \#_1(a) \}}$

\end{lstlisting}
\end{minipage}

\caption{Exemple derivation using amortized reasoning, we noted
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$, we use the
  tick metric that assigns cost $n$ to \lstinline{tick(n)} statements and
  0 to other operations.
  }
\label{fig:xmplinc}
\end{figure}

\begin{figure}
\begin{lstlisting}
${\color{blue}%
  \{ (Z = \log_2(h-l)) + Z \} }$
bsearch(x,l,h) {
  if (h-l > 1) {
    ${\color{blue}%
      \{ (h-l > 1 \land Z = \log_2(h-l)) + Z \} }$
    m = h + (h-l)/2;
    ${\color{blue}%
      \{ (m = \frac{h+l}{2} \land h-l > 1 \land Z = \log_2(h-l)) + Z \} }$
    if (a[m]>x) h=m; else l=m;
    ${\color{blue}%
      \{ (Z - 1 = \log_2(h-l)) + (Z-1) + 1 \} }$
    l = bsearch(x,l,h); }
  ${\color{blue}%
    \{ Z-1 \} }$
  return l;
} ${\color{blue} \{ Z \} }$
\end{lstlisting}
\caption{Exemple derivation of a stack usage bound for a binary
  search program, the metric used assigns a cost of 1 to function
  calls and $-1$ to return statements, we noted $Z$ an arbitrary
  logical variable.
  }
\end{figure}

@@

\and \Rule{Q:Set}
{ \forall u.\, q_{yu} = q'_{xu} + q'_{yu}
\\ q'_{xy} \in \mathbb Q_0^+
}
{ \Gamma[x/y]; Q + M_s \vdash x \gets y \dashv Q'; \Gamma }

@@

\and \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |x - y| \}
\\ \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |x - y| \le u_{xy} \}
%\\ q'_0 = q_0 + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i
\\ \forall i \in \mathcal U.\, q'_i \ge q_i - p_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U \cup \mathcal L.\, q'_i \ge q_i
}
{ Q' + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i  \succeq_\Gamma Q }

@@

The rules of~\pref{fig:proof} are designed to enforce a semantic
soundness similar to the one exposed in the overview section.  To
do this, the rules impose contraints on $Q$ and $Q'$.  To make the
presentation lighter, we use the following  set of conventions.
First, the context $Q'$ noted $Q \pm n$ is defined by $q'_0 = q_0 \pm n$
and $\forall i \neq 0.\, q'_i = q_i$.  Second, for any index $i$ such that
the relation between $q_i$ and $q'_i$ is not specified as a side condition
of a rule, we assume that $q_i = q'_i$.  Third, we implicitely assume
all the side conditions enforcing that the coefficients $(q_i)_{i \in I}$ are
non-negative numbers. Finally, when we write $q_x$ for $q_{x0}$ where
0 is a program variable symbolizing the constant 0.

@@

Syntactically $\Gamma$ is a list of program conditions.  Because of the
rules {\sc Q:While} and {\sc Q:If}, we require the program conditions
to be closed by negation.  Semantically we define the validity
$H \models \Gamma$ of $\Gamma = C_1, \dots, C_n$ on a given
heap $H$ to be $\bigwedge_i \llbracket C_i \rrbracket_H = \top$.
%
The logical state
is maintained using regular Hoare style rules.
The semantic entailment on logical states is noted $\Gamma \models \Delta$,
its meaning is that $\forall H.\, H\models\Gamma \implies H\models\Delta$.

@@

\begin{lemma}[Relax]
If $Q' \succeq_\Gamma Q$, for any heap $H$ such that $H \models \Gamma$,
we have $\Phi_H(Q') \ge \Phi_H(Q)$.
\end{lemma}
\begin{proof}
We observe the following inequations.
\begin{align*}
\Phi_H(Q') &= \sum_i q'_i f_i(H) \\
&\ge q_0 + \sum_{i\in\mathcal U} u_i p_i - \sum_{i\in\mathcal L} l_i p_i
  + \sum_{i\in\mathcal U} (q_i - p_i) f_i(H) + \sum_{i\in\mathcal L} (q_i + p_i) f_i(H)
  + \sum_{i\not\in \mathcal U \cup \mathcal L} q_i \\
 &\ge \Phi_H(Q)
  + \sum_{i\in\mathcal U} (u_i - f_i(H)) p_i
  + \sum_{i\in\mathcal L} (f_i(H) - l_i) p_i \\
 &\ge \Phi_H(Q).
\end{align*}
Since $H \models \Gamma$, using the transitivity of $\models$ and
the definition of $\mathcal U$ and $\mathcal L$ we get
$\forall i\in\mathcal L.\, l_i \le f_i(H)$ and
$\forall i\in\mathcal U.\, f_i(H)\le u_i$.
These two inequalities justify the last step.
\end{proof}

@@

\item Case {\sc S:Seq}.
  By transitivity and induction hypothesis, it is
  easy to get $H'' \models \Gamma''$.  Now we prove
  the quantitative part of the theorem.  We have
  $(q, q') = M_s^1 \cdot (q_1, q'_1) \cdot M_s^2 \cdot (q_2, q'_2)$.
  We assume $\forall i \in \{1, 2\}.\, M_s^i \ge 0$, the other cases
  are similar.
  \begin{itemize}[topsep=0pt]
  \item
    The first case is when $q'_1 > M_s^2 + q_2$ then $q = M_s^1 + q_1$
    and since $q_1 \le \Phi_H(Q - M_s^1)$ (by induction) we have
    $q \le \Phi_H(Q)$.
  \item
    The second case is when $q'_1 \le M_s^2 + q_2$ then $q = M_s^1 + q_1 + M_s^2 + q_2 - q'_1$. \\
    Since $\Phi_H(Q - M_s^1) - \Phi_{H'}(Q') \ge q_1 - q'_1$
    and $q_2 \le \Phi_{H'}(Q' - M_s^2)$ we have:
    \begin{align*}
    q &\le M_s^1 + \Phi_H(Q - M_s^1) - \Phi_{H'}(Q') + M_s^2 + q_2 \\
      &= \Phi_H(Q) + (q_2 - \Phi_{H'}(Q' - M_s^2)) \le \Phi_H(Q).
   \end{align*}
  \end{itemize}
  We obtain $\Phi_H(Q) - \Phi_{H''}(Q'') \ge q - q'$ by simply chaining
  the two induction hypothesis and using the fact that $q - q' =
  (q_1 + M_s^1 - q'_1) + (q_2 + M_s^2 - q'_2)$.

@@

\item Case {\sc S:WhileStep}.
  Here again we assume $\forall i \in \{1, 2, 3\}.\, M_w^i \ge 0$.
  \begin{itemize}[topsep=0pt]
  \item
    If $q'_1 > M_w^2 + q_2$ then $q = M_w^1 + q_1$ and by induction
    we get $q_1 \le \Phi_H(Q - M_w^1)$, so $q \le \Phi_H(Q)$.
  \item
    If $q'_1 \le M_w^2 + q_2$ then $q = M_w^1 + q_1 + M_w^2 + q_2 - q'_1$.
    And we proceed like in the sequence case.
  \end{itemize}
  We obtain $\Phi_H(Q) - \Phi_{H''}(Q - M_w^3) \ge q - q'$ using the identity
  $q - q' = (M_w^1 + M_w^2 + q_1 - q'_1) + q_2 - q'_2$ and the two
  induction hypothesis
  \begin{itemize}[topsep=0pt]
  \item $\Phi_H(Q - M_w^1) - \Phi_{H'}(Q + M_w^2) \ge q_1 - q'_1$ and
  \item $\Phi_{H'}(Q) - \Phi_{H''}(Q - M_w^3) \ge q_2 - q'_2.$
  \end{itemize}

@@

\paragraph{Decision procedure.}
To decide logical entailment $\Gamma \models \Delta$ we use
Presburger's decision procedure.  But what this procedure gives us
is in fact only an answer the the question: Is there any heap $H$
such that $H \models \Gamma$.  While this sentence is existential
the question $\Gamma \models \Delta$ is universal.  To use the
decision procedure, we use the well known trick of double negation.
\begin{align*}
\Gamma \models C
&\Leftrightarrow \neg\neg (\Gamma \models C) \\
&\Leftrightarrow \neg\neg (\forall H.\, H\models\Gamma \implies H\models C) \\
&\Leftrightarrow \neg \exists H, H\models\Gamma \land \neg H\models C \\
&\Leftrightarrow \neg \exists H, H\models\Gamma \land H\models \neg C \\
&\Leftrightarrow \neg \exists H, H\models (\Gamma, \neg C)
\end{align*}
Thus, to know if $\Gamma \models C$ we ask the decision
procedure if $\Gamma, \neg C$ is satisfiable and negate the result.
When we want to know about $\Gamma \models \Delta$ we simply ask
one question for each of the conjuncts of $\Delta$.
